---
title: "Now Departing: Three Quarters of Every Dollar Earned in Hancock Renovations Left Onondaga County"
author: "Jamison R. Crawford, MPA"
date: "May 3, 2019"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: true
---

<style>
body {
text-align: justify}
</style>

# Overview

The following outlines the exodus of nearly 75 cents on the dollar from Onondaga County by itinerate workers imported from other counties for the 2018 Hancock Airport renovations. These findings are completely reproducible from the raw data, provided by Syracuse Regional Airport Authority, and may be found on my [GitHub](https://github.com/jamisoncrawford/wealth).

This is a slapdash publication, which will be formatted with human-readable text this weekend.

# Load Packages

```{r echo=TRUE, message=FALSE, warning=FALSE}
if(!require("readr")){install.packages("readr")}
if(!require("dplyr")){install.packages("dplyr")}
if(!require("tigris")){install.packages("tigris")}
if(!require("scales")){install.packages("scales")}
if(!require("stringr")){install.packages("stringr")}
if(!require("ggplot2")){install.packages("ggplot2")}
if(!require("noncensus")){install.packages("noncensus")}
if(!require("kableExtra")){install.packages("kableExtra")}

library(readr)
library(dplyr)
library(tigris)
library(scales)
library(stringr)
library(ggplot2)
library(noncensus)
library(kableExtra)
```

# Read in Data

```{r}
data(zip_codes)
data(counties)

url <- "https://raw.githubusercontent.com/jamisoncrawford/wealth/master/Tidy%20Data/hancock_lakeview_tidy.csv"
lvhc <- read_csv(file = url, 
                 col_types = "ccDcccddddccliii")

rm(url)
```

# Convert ZIP to Counties

```{r}

counties <- counties %>%
  mutate(fips = paste0(state_fips, 
                       county_fips),                        # Paste FIPS IDs (State + County)
         fips = as.character(fips)) %>%
  select(county_name, fips) 

lvhc <- lvhc %>%
  filter(!is.na(zip),
         !is.na(gross),
         !is.na(project)) %>%                               # Remove NA values
  left_join(zip_codes, 
            by = "zip") %>%
  mutate(fips = as.character(fips)) %>%                     # Join FIPS IDs
  left_join(counties,
            by = "fips") %>%                                # Join counties by FIPS
  mutate(county = str_replace_all(string = county_name, 
                                  pattern = " County$", 
                                  replacement = "")) %>%    # Clean county names
  select(project:hours, gross, sex:race, city:state, 
         county, fips, longitude:latitude) %>%              # Select variables
  filter(!is.na(state))                                     # Remove 8 unidentified records
  
rm(zip_codes, counties)

lv <- lvhc %>%
  filter(project == "Lakeview")

hc <- lvhc %>%
  filter(project == "Hancock")

```

# Summaries by Group

```{r}

hc_gross <- hc %>%
  group_by(county, state) %>%
  summarize(gross = sum(gross)) %>%       # Total gross by county
  ungroup() %>%
  arrange(desc(gross)) %>%
  mutate(perc_gross = gross / sum(gross)) # Percentage of total gross

hc_inds <- hc %>%
  group_by(name, zip, ssn, class, sex, race) %>%
  summarize(gross = sum(gross),
            county = unique(county),
            state = unique(state)) %>%    # Total gross by individual
  ungroup() %>%
  arrange(county, desc(gross))

hc_wrks <- hc_inds %>%
  group_by(county, state) %>%
  summarize(workers = n(),
            gross = sum(gross)) %>%
  ungroup() %>%
  arrange(desc(gross))                    # Total workers by county

hc_all <- hc_gross %>%
  left_join(hc_wrks,
            by = c("county", "state")) %>%
  select(-gross.y) %>%
  rename(gross = gross.x)                 # Merge workers and gross

hc_all <- hc_all %>%
  mutate(perc_workers = workers / sum(workers))

rm(hc_gross, hc_wrks)


```

# Visualizing an Exodus

```{r}

options(scipen = 999)                     # Disable scientific notation

ggplot(hc_all, 
       aes(x = reorder(county, gross), gross, 
           y = gross, 
           fill = state)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Total gross income by county, state",
       subtitle = "Hancock Airport Renovations",
       caption = "Source: Syracuse Regional Airport Authority",
       x = "County",
       y = "Total Gross (USD)",
       fill = "State") +
  scale_y_continuous(labels = comma) +
  theme_minimal()

```

# Summary Table

```{r}
hc_all_tbl <- hc_all %>%
  mutate(gross = dollar(gross),
         perc_gross = percent(perc_gross),
         perc_workers = percent(perc_workers)) %>%
  rename(County = county,
         State = state,
         "Total Gross" = gross,
         "Gross (%)" = perc_gross,
         "Total Workers" = workers,
         "Workforce (%)" = perc_workers)

kable(hc_all_tbl) %>%
  kable_styling(bootstrap_options = c("striped", 
                                      "hover", 
                                      "responsive"),
                full_width = TRUE)
```
